Component response rate variation underlies stability in large complex systems

### 2019-02-22 notebook/ms.tex
Add some to the abstract to emphasise finite systems

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{setspace}
\setstretch{2}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{unicode=true,
            pdftitle={Component response rate variation drives stability in large complex systems},
            pdfauthor={A. Bradley Duthie ( alexander.duthie@stir.ac.uk )},
            colorlinks=true,
            linkcolor=blue,
            citecolor=Blue,
            urlcolor=Blue,
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Component response rate variation drives stability in large complex
systems}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{A. Bradley Duthie (
\href{mailto:alexander.duthie@stir.ac.uk}{\nolinkurl{alexander.duthie@stir.ac.uk}}
)}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{Biological and Environmental Sciences, University of Stirling, Stirling,
UK, FK9 4LA}

@@

The stability of a complex system generally decreases with increasing
system size and interconnectivity, a counterintuitive result of
widespread importance across the physical, life, and social sciences.
Despite recent interest in the relationship between system properties
and stability, the effect of variation in the response rate of
individual system components remains unconsidered. Here I vary the
component response rates (\(\boldsymbol{\gamma}\)) of randomly generated
complex systems. I show that when component response rates vary, the
potential for system stability is markedly increased. Variation in
\(\boldsymbol{\gamma}\) becomes increasingly important as system size
increases, such that the largest stable complex systems would be
unstable if not for \(\boldsymbol{Var(\gamma)}\). My results reveal a
previously unconsidered driver of system stability that is likely to be
pervasive across all complex systems.

@@

In 1972, May {[}\protect\hyperlink{ref-May1972}{1}{]} first demonstrated
that randomly assembled systems of sufficient complexity are almost
inevitably unstable given infinitesimally small perturbations.
Complexity in this case is defined by the size of the system (i.e., the
number of potentially interacting components; \(S\)), its connectance
(i.e., the probability that one component will interact with another;
\(C\)), and the variance of interaction strengths (\(\sigma^{2}\))
{[}\protect\hyperlink{ref-Allesina2012}{2}{]}. May's finding that the
probability of local stability falls to near zero given a sufficiently
high threshold of \(\sigma\sqrt{SC}\) is broadly relevant for
understanding the dynamics and persistence of systems such as ecological
{[}\protect\hyperlink{ref-May1972}{1}--\protect\hyperlink{ref-Grilli2017}{5}{]},
neurological
{[}\protect\hyperlink{ref-Gray2008}{6},\protect\hyperlink{ref-Gray2009}{7}{]},
biochemical
{[}\protect\hyperlink{ref-Rosenfeld2009}{8},\protect\hyperlink{ref-MacArthur2010}{9}{]},
and socio-economic
{[}\protect\hyperlink{ref-May2008}{10}--\protect\hyperlink{ref-Bardoscia2017}{13}{]}
networks. As such, identifying general principles that drive stability
in complex systems is of wide-ranging importance.

Randomly assembled complex systems can be represented as large square
matrices (\(\mathbf{M}\)) with \(S\) components (e.g., networks of
species {[}\protect\hyperlink{ref-Allesina2012}{2}{]} or banks
{[}\protect\hyperlink{ref-Haldane2011}{11}{]}). One element of such a
matrix, \(M_{ij}\), defines how component \(j\) affects component \(i\)
in the system at a point of equilibrium
{[}\protect\hyperlink{ref-Allesina2012}{2}{]}. Off-diagonal elements
(\(i \neq j\)) therefore define interactions between components, while
diagonal elements (\(i = j\)) define component self-regulation (e.g.,
carrying capacity in ecological communities). Traditionally,
off-diagonal elements are assigned non-zero values with a probability
\(C\), which are sampled from a distribution with variance
\(\sigma^{2}\); diagonal elements are set to -1
{[}\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{4}{]}.
Local system stability is assessed using eigenanalysis, with the system
being stable if the real parts of all eigenvalues (\(\lambda\)) of
\(\mathbf{M}\) are negative (\(\max\left(\Re(\lambda)\right) < 0\))
{[}\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}{]}.
In a large system (high \(S\)), eigenvalues are distributed uniformly
{[}\protect\hyperlink{ref-Tao2010}{14}{]} within a circle centred at
\(\Re = -1\) (the mean value of diagonal elements) and \(\Im = 0\), with
a radius of \(\sigma\sqrt{SC}\)
{[}\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{4}{]}
(Figs 1a and 2a). Local stability of randomly assembled systems
therefore becomes increasingly unlikely as \(S\), \(C\), and
\(\sigma^{2}\) increase.

May's
{[}\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}{]}
stability criterion \(\sigma\sqrt{SC} < 1\) assumes that individual
components respond to perturbations of the system at the same rate
(\(\gamma\)), but this is highly unlikely in any complex system. In
ecological communities, for example, the rate at which population
density changes following perturbation will depend on the generation
time of organisms, which might vary by orders of magnitude among
species. Species with short generation times will respond quickly (high
\(\gamma\)) to perturbations relative to species with long generation
times (low \(\gamma\)). Similarly, the speed at which individual banks
respond to perturbations in financial networks, or individuals or
institutions respond to perturbations in complex social networks, is
likely to vary. The effect of such variance on stability has not been
investigated in complex systems theory. Intuitively, variation in
\(\gamma\) might be expected to decrease system stability by introducing
a new source of variation into the system and thereby increasing
\(\sigma\). Here I show why, despite higher \(\sigma\), complex systems
in which \(\gamma\) varies are actually more likely to be stable,
especially when \(S\) is high.

@@

\textbf{Component response rates of random complex systems}. Rows in
\(\mathbf{M}\) define how a given component \(i\) is affected by other
components of the system, meaning that the rate of component response
time can be modelled by multiplying all row elements by a real scalar
value \(\gamma_{i}\) {[}\protect\hyperlink{ref-Patel2018}{15}{]}. The
distribution of \(\gamma\) over \(S\) components thereby models the
distribution of component response rates. An instructive example
compares one \(\mathbf{M}\) where \(\gamma_{i} = 1\) for all \(i\) in
\(S\) to the same \(\mathbf{M}\) when half of \(\gamma_{i} = 1.95\) and
half of \(\gamma_{i} = 0.05\). This models one system in which
\(\gamma\) is invariant and one in which \(\gamma\) varies, but systems
are otherwise identical (note \(E[\gamma_{i}] = 1\) in both cases). I
assume \(S = 200\), \(C = 0.05\), and \(\sigma = 0.4\); diagonal
elements are set to \(-1\) and non-zero off-diagonal elements are drawn
randomly from \(\mathcal{N}(0, \sigma^{2})\). Rows are then multiplied
by \(\gamma_{i}\) to generate \(\mathbf{M}\). When \(\gamma_{i} = 1\),
eigenvalues of \(\mathbf{M}\) are distributed uniformly within a circle
centred at (\(-1, 0\)) with a radius of 1.265 (Fig. 1a). Hence, the real
components of eigenvalues are highly unlikely to all be negative when
all \(\gamma_{i} = 1\). But when \(\gamma_{i}\) values are separated
into two groups, eigenvalues are no longer uniformly distributed (Fig.
1b). Instead, two distinct clusters of eigenvalues appear (red circles
in Fig. 1b), one centred at (\(-1.95, 0\)) and the other centred at
(\(-0.05, 0\)). The former has a large radius, but the real components
have shifted to the left (in comparison to when \(\gamma = 1\)), while
the latter cluster has a smaller radius, but real components have
shifted to the right; all \(\Re({\lambda}) < 0\). Overall, for 1 million
randomly assembled \(\mathbf{M}\), this division between slow and fast
component response rates results in more stable systems: 1 stable given
\(\gamma = 1\) versus 32 stable given \(\gamma = \{1.95, 0.5\}\).

Higher stability in systems with variation in \(\gamma\) can be observed
by sampling \(\gamma_{i}\) values from various distributions. I focus on
a uniform distribution where \(\gamma \sim \mathcal{U}(0, 2)\) (see
Supplementary Information for other distributions, which give similar
results). As with the case of \(\gamma = \{1.95, 0.5\}\) (Fig. 1b),
\(E[\gamma] = 1\) when \(\gamma \sim \mathcal{U}(0, 2)\), allowing
comparison of \(\mathbf{M}\) before and after the addition of variation
in component response rate. Figure 2 shows a comparison of eigenvalue
distributions given \(S = 1000\), \(C = 0.05\), and \(\sigma = 0.4\). As
expected {[}\protect\hyperlink{ref-Tao2010}{14}{]}, when \(\gamma = 1\),
eigenvalues are distributed uniformly in a circle centred at (\(-1, 0\))
with a radius of \(\sigma\sqrt{SC} =\) 2.828. Uniform variation in
\(\gamma\) leads to a non-uniform distribution of eigenvalues
{[}\protect\hyperlink{ref-Ahmadian2015}{16}--\protect\hyperlink{ref-Stone2017}{18}{]},
some of which are clustered locally near the centre of the distribution,
but others of which are spread outside the former radius of 2.828 (Fig
2b). The clustering and spreading of eigenvalues introduced by
\(Var(\gamma)\) can destabilise previously stable systems or stabilise
systems that are otherwise unstable. But where systems are otherwise too
complex to be stable given \(\gamma = 1\), the effect of \(Var(\gamma)\)
can often lead to stability above May's
{[}\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}{]}
threshold \(\sigma\sqrt{SC} < 1\).

@@

\textbf{Targeted manipulation of \(\mathbf{\gamma}\)}. To further
investigate the potential of \(Var(\gamma)\) to be stabilising, I used a
genetic algorithm. Genetic algorithms are heuristic tools that mimic
evolution by natural selection, and are useful when the space of
potential solutions (in this case, possible combinations of \(\gamma\)
values leading to stability in a large complex system) is too large to
search exhaustively {[}\protect\hyperlink{ref-Hamblin2013}{19}{]}.
Generations of selection on \(\gamma\) value combinations to minimise
\(\max\left(\Re(\lambda)\right)\) demonstrated the potential for
\(Var(\gamma)\) to increase system stability. Across
\(S = \{2, 3, ..., 39, 40\}\), sets of \(\gamma\) values were found that
resulted in stable systems with probabilities that were up to four
orders of magnitude higher than when \(\gamma = 1\) (Fig. 4), meaning
that stability could often be achieved by manipulating \(S\) \(\gamma\)
values rather than \(S \times S\) \(\mathbf{M}\) elements.

\textbf{System feasibility given \(\mathbf{Var(\gamma)}\)} For complex
systems in which individual system components represent the density of
some tangible quantity, it is relevant to consider the feasibility of
the system. Feasibilility assumes that the system is stable and values
of all components are positive at equilibrium
{[}\protect\hyperlink{ref-Grilli2017}{5},\protect\hyperlink{ref-Dougoud2018}{20},\protect\hyperlink{ref-Song2018}{21}{]}.
This is of particular interest for ecological communities because
population density cannot take negative values, meaning that ecological
systems need to be feasible for stability to be biologically realistic
{[}\protect\hyperlink{ref-Dougoud2018}{20}{]}. While my results are
intended to be general to all complex systems, and not restricted to
species networks, I have also performed a feasibility analysis on all
matrices \(\mathbf{M}\) tested for stability, and additionally for
specific types of ecological communities
{[}\protect\hyperlink{ref-Allesina2012}{2}{]} (e.g., competitive,
mutualist, predator-prey; see Supplementary Information). Feasibility
was unaffected by \(Var(\gamma)\) and instead occurred with a fixed
probability of \(1/2^{S}\), consistent with a recent proof by Serván et
al. {[}\protect\hyperlink{ref-Servan2018}{22}{]} (see Supplementary
Information). Hence, for pure interacting species networks, variation in
component response rate (i.e., species generation time) does not affect
stability at biologically realistic species densities.

@@

Here I have shown that the stability of large systems might often be
contigent upon variation in the response rates of their individual
components, meaning that factors such as rate of trait evolution (in
biological networks), transaction speed (in economic networks), or
communication speed (in social networks) need to be considered when
investigating the stability of complex systems. Variation in component
response rate becomes more likely to be critical for stability as system
size increases, and can ultimately increase the overall probability that
system stability is observed above that predicted by May's
{[}\protect\hyperlink{ref-May1972}{1}{]} classically derived
\(\sigma \sqrt{SC}\) criterion. The logic outlined here is general, and
potentially applies to any complex system in which individual system
components can vary in their reaction rates to system perturbation.

It is important to emphasise that variation in component response rate
is not stabilising per se; that is, adding variation in component
response rate to a particular system does not necessarily increase the
probability that the system will be stable. Rather, systems that are
observed to be stable are more likely to have varying component response
rates, and for this variation to be critical to their stability (Fig.
3). This is caused by the shift to a non-uniform distribution of
eigenvalues that occurs by introducing \(Var(\gamma)\) (Fig. 1b, 2b),
which can sometimes cause all of the real components of the eigenvalues
of the community matrix \(\mathbf{M}\) to become negative, but might
also increase the real components of eigenvalues. The mathematics
underlying this shift in eigenvalue distribution has been investigated
{[}\protect\hyperlink{ref-Ahmadian2015}{16}{]} and recently applied to
questions concerning species density and feasibility
{[}\protect\hyperlink{ref-Gibbs2017}{17},\protect\hyperlink{ref-Stone2017}{18}{]},
but has not been interpreted as rates of response of individual system
components to perturbation.

@@

Interestingly, while complex systems were more likely to be stable given
variation in component response rate, they were not more likely to be
feasible, meaning that stability was not increased when component values
were also restricted to being positive at equilibrium. Feasibility is
important to consider, particularly for the study of ecological networks
of species
{[}\protect\hyperlink{ref-Grilli2017}{5},\protect\hyperlink{ref-Stone2017}{18},\protect\hyperlink{ref-Dougoud2018}{20},\protect\hyperlink{ref-Servan2018}{22}{]}
because population densities cannot realistically be negative. My
results therefore suggest that variation in the rate of population
responses to perturbation (e.g., due to differences in generation time
among species) is unlikely to be critical to the feasibility of purely
multi-species interaction networks (see also Supplementary Information).
Nevertheless, ecological interactions do not exist in isolation in
empirical systems {[}\protect\hyperlink{ref-Patel2018}{15}{]}, but
instead interact with evolutionary, abiotic, or social-economic systems.
The relevance of component response rate for complex system stability
should therefore not be ignored in the broader context of ecological
communities.

@@

\textbf{Component response rate variation (\(\mathbf{\gamma}\))}. In a
synthesis of eco-evolutionary feedbacks on community stability, Patel et
al. model a system that includes a vector of potentially changing
species densities (\(\mathbf{N}\)) and a vector of potentially evolving
traits (\(\mathbf{x}\)) {[}\protect\hyperlink{ref-Patel2018}{15}{]}. For
any species \(i\) or trait \(j\), change in species density (\(N_{i}\))
or trait value (\(x_{j}\)) with time (\(t\)) is a function of the
vectors \(\mathbf{N}\) and \(\mathbf{x}\),

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{setspace}
\setstretch{2}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{unicode=true,
            pdftitle={Component response rate variation underlies stability in large complex systems},
            pdfauthor={A. Bradley Duthie ( alexander.duthie@stir.ac.uk )},
            colorlinks=true,
            linkcolor=blue,
            citecolor=Blue,
            urlcolor=Blue,
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}

  \title{Component response rate variation underlies stability in large complex
systems}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{A. Bradley Duthie (
\href{mailto:alexander.duthie@stir.ac.uk}{\nolinkurl{alexander.duthie@stir.ac.uk}}
)}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{Biological and Environmental Sciences, University of Stirling, Stirling,
UK, FK9 4LA}

@@

The stability of a complex system generally decreases with increasing
system size and interconnectivity, a counterintuitive result of
widespread importance across the physical, life, and social sciences.
Despite recent interest in the relationship between system properties
and stability, the effect of variation in the response rate of
individual system components remains unconsidered. Here I vary the
component response rates (\(\boldsymbol{\gamma}\)) of randomly generated
complex systems. I show that when component response rates vary, the
potential for system stability is markedly increased. Variation in
\(\boldsymbol{\gamma}\) becomes increasingly important as system size
increases, such that the largest stable complex systems would be
unstable if not for \(\boldsymbol{Var(\gamma)}\). My results reveal a
previously unconsidered aspect of system stability that is likely to be
pervasive across all complex systems.

@@

In 1972, May\textsuperscript{\protect\hyperlink{ref-May1972}{1}} first
demonstrated that randomly assembled systems of sufficient complexity
are almost inevitably unstable given infinitesimally small
perturbations. Complexity in this case is defined by the size of the
system (i.e., the number of potentially interacting components; \(S\)),
its connectance (i.e., the probability that one component will interact
with another; \(C\)), and the variance of interaction strengths
(\(\sigma^{2}\))\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}.
May's finding that the probability of local stability falls to near zero
given a sufficiently high threshold of \(\sigma\sqrt{SC}\) is broadly
relevant for understanding the dynamics and persistence of systems such
as
ecological\textsuperscript{\protect\hyperlink{ref-May1972}{1}--\protect\hyperlink{ref-Grilli2017}{5}},
neurological\textsuperscript{\protect\hyperlink{ref-Gray2008}{6},\protect\hyperlink{ref-Gray2009}{7}},
biochemical\textsuperscript{\protect\hyperlink{ref-Rosenfeld2009}{8},\protect\hyperlink{ref-MacArthur2010}{9}},
and
socio-economic\textsuperscript{\protect\hyperlink{ref-May2008}{10}--\protect\hyperlink{ref-Bardoscia2017}{13}}
networks. As such, identifying general principles that affect stability
in complex systems is of wide-ranging importance.

Randomly assembled complex systems can be represented as large square
matrices (\(\mathbf{M}\)) with \(S\) components (e.g., networks of
species\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}} or
banks\textsuperscript{\protect\hyperlink{ref-Haldane2011}{11}}). One
element of such a matrix, \(M_{ij}\), defines how component \(j\)
affects component \(i\) in the system at a point of
equilibrium\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}.
Off-diagonal elements (\(i \neq j\)) therefore define interactions
between components, while diagonal elements (\(i = j\)) define component
self-regulation (e.g., carrying capacity in ecological communities).
Traditionally, off-diagonal elements are assigned non-zero values with a
probability \(C\), which are sampled from a distribution with variance
\(\sigma^{2}\); diagonal elements are set to
-1\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{4}}.
Local system stability is assessed using eigenanalysis, with the system
being stable if the real parts of all eigenvalues (\(\lambda\)) of
\(\mathbf{M}\) are negative
(\(\max\left(\Re(\lambda)\right) < 0\))\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}.
In a large system (high \(S\)), eigenvalues are distributed
uniformly\textsuperscript{\protect\hyperlink{ref-Tao2010}{14}} within a
circle centred at \(\Re = -1\) (the mean value of diagonal elements) and
\(\Im = 0\), with a radius of
\(\sigma\sqrt{SC}\)\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{4}}
(Figs 1a and 2a). Local stability of randomly assembled systems
therefore becomes increasingly unlikely as \(S\), \(C\), and
\(\sigma^{2}\) increase.

May's\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}
stability criterion \(\sigma\sqrt{SC} < 1\) assumes that individual
components respond to perturbations of the system at the same rate
(\(\gamma\)), but this is highly unlikely in any complex system. In
ecological communities, for example, the rate at which population
density changes following perturbation will depend on the generation
time of organisms, which might vary by orders of magnitude among
species. Species with short generation times will respond quickly (high
\(\gamma\)) to perturbations relative to species with long generation
times (low \(\gamma\)). Similarly, the speed at which individual banks
respond to perturbations in financial networks, or individuals or
institutions respond to perturbations in complex social networks, is
likely to vary. The effect of such variance on stability has not been
investigated in complex systems theory. Intuitively, variation in
\(\gamma\) might be expected to decrease system stability by introducing
a new source of variation into the system and thereby increasing
\(\sigma\). Here I show why, despite higher \(\sigma\), complex systems
in which \(\gamma\) varies are actually more likely to be stable,
especially when \(S\) is high.

@@

\textbf{Component response rates of random complex systems}. Rows in
\(\mathbf{M}\) define how a given component \(i\) is affected by other
components of the system, meaning that the rate of component response
time can be modelled by multiplying all row elements by a real scalar
value
\(\gamma_{i}\)\textsuperscript{\protect\hyperlink{ref-Patel2018}{15}}.
The distribution of \(\gamma\) over \(S\) components thereby models the
distribution of component response rates. An instructive example
compares one \(\mathbf{M}\) where \(\gamma_{i} = 1\) for all \(i\) in
\(S\) to the same \(\mathbf{M}\) when half of \(\gamma_{i} = 1.95\) and
half of \(\gamma_{i} = 0.05\). This models one system in which
\(\gamma\) is invariant and one in which \(\gamma\) varies, but systems
are otherwise identical (note \(E[\gamma_{i}] = 1\) in both cases). I
assume \(S = 200\), \(C = 0.05\), and \(\sigma = 0.4\); diagonal
elements are set to \(-1\) and non-zero off-diagonal elements are drawn
randomly from \(\mathcal{N}(0, \sigma^{2})\). Rows are then multiplied
by \(\gamma_{i}\) to generate \(\mathbf{M}\). When \(\gamma_{i} = 1\),
eigenvalues of \(\mathbf{M}\) are distributed uniformly within a circle
centred at (\(-1, 0\)) with a radius of 1.265 (Fig. 1a). Hence, the real
components of eigenvalues are highly unlikely to all be negative when
all \(\gamma_{i} = 1\). But when \(\gamma_{i}\) values are separated
into two groups, eigenvalues are no longer uniformly distributed (Fig.
1b). Instead, two distinct clusters of eigenvalues appear (red circles
in Fig. 1b), one centred at (\(-1.95, 0\)) and the other centred at
(\(-0.05, 0\)). The former has a large radius, but the real components
have shifted to the left (in comparison to when \(\gamma = 1\)), while
the latter cluster has a smaller radius, but real components have
shifted to the right; all \(\Re({\lambda}) < 0\). Overall, for 1 million
randomly assembled \(\mathbf{M}\), this division between slow and fast
component response rates results in more stable systems: 1 stable given
\(\gamma = 1\) versus 32 stable given \(\gamma = \{1.95, 0.5\}\).

Higher stability in systems with variation in \(\gamma\) can be observed
by sampling \(\gamma_{i}\) values from various distributions. I focus on
a uniform distribution where \(\gamma \sim \mathcal{U}(0, 2)\) (see
Supplementary Information for other distributions, which give similar
results). As with the case of \(\gamma = \{1.95, 0.5\}\) (Fig. 1b),
\(E[\gamma] = 1\) when \(\gamma \sim \mathcal{U}(0, 2)\), allowing
comparison of \(\mathbf{M}\) before and after the addition of variation
in component response rate. Figure 2 shows a comparison of eigenvalue
distributions given \(S = 1000\), \(C = 0.05\), and \(\sigma = 0.4\). As
expected\textsuperscript{\protect\hyperlink{ref-Tao2010}{14}}, when
\(\gamma = 1\), eigenvalues are distributed uniformly in a circle
centred at (\(-1, 0\)) with a radius of \(\sigma\sqrt{SC} =\) 2.828.
Uniform variation in \(\gamma\) leads to a non-uniform distribution of
eigenvalues\textsuperscript{\protect\hyperlink{ref-Ahmadian2015}{16}--\protect\hyperlink{ref-Stone2017}{18}},
some of which are clustered locally near the centre of the distribution,
but others of which are spread outside the former radius of 2.828 (Fig
2b). The clustering and spreading of eigenvalues introduced by
\(Var(\gamma)\) can destabilise previously stable systems or stabilise
systems that are otherwise unstable. But where systems are otherwise too
complex to be stable given \(\gamma = 1\), the effect of \(Var(\gamma)\)
can often lead to stability above
May's\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}
threshold \(\sigma\sqrt{SC} < 1\).

@@

\textbf{Targeted manipulation of \(\mathbf{\gamma}\)}. To further
investigate the potential of \(Var(\gamma)\) to be stabilising, I used a
genetic algorithm. Genetic algorithms are heuristic tools that mimic
evolution by natural selection, and are useful when the space of
potential solutions (in this case, possible combinations of \(\gamma\)
values leading to stability in a large complex system) is too large to
search
exhaustively\textsuperscript{\protect\hyperlink{ref-Hamblin2013}{19}}.
Generations of selection on \(\gamma\) value combinations to minimise
\(\max\left(\Re(\lambda)\right)\) demonstrated the potential for
\(Var(\gamma)\) to increase system stability. Across
\(S = \{2, 3, ..., 39, 40\}\), sets of \(\gamma\) values were found that
resulted in stable systems with probabilities that were up to four
orders of magnitude higher than when \(\gamma = 1\) (Fig. 4), meaning
that stability could often be achieved by manipulating \(S\) \(\gamma\)
values rather than \(S \times S\) \(\mathbf{M}\) elements.

\textbf{System feasibility given \(\mathbf{Var(\gamma)}\)} For complex
systems in which individual system components represent the density of
some tangible quantity, it is relevant to consider the feasibility of
the system. Feasibilility assumes that the system is stable and values
of all components are positive at
equilibrium\textsuperscript{\protect\hyperlink{ref-Grilli2017}{5},\protect\hyperlink{ref-Dougoud2018}{20},\protect\hyperlink{ref-Song2018}{21}}.
This is of particular interest for ecological communities because
population density cannot take negative values, meaning that ecological
systems need to be feasible for stability to be biologically
realistic\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{20}}.
While my results are intended to be general to all complex systems, and
not restricted to species networks, I have also performed a feasibility
analysis on all matrices \(\mathbf{M}\) tested for stability, and
additionally for specific types of ecological
communities\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}
(e.g., competitive, mutualist, predator-prey; see Supplementary
Information). Feasibility was unaffected by \(Var(\gamma)\) and instead
occurred with a fixed probability of \(1/2^{S}\), consistent with a
recent proof by Serván et
al.\textsuperscript{\protect\hyperlink{ref-Servan2018}{22}} (see
Supplementary Information). Hence, for pure interacting species
networks, variation in component response rate (i.e., species generation
time) does not affect stability at biologically realistic species
densities.

@@

Here I have shown that the stability of large systems might often be
contigent upon variation in the response rates of their individual
components, meaning that factors such as rate of trait evolution (in
biological networks), transaction speed (in economic networks), or
communication speed (in social networks) need to be considered when
investigating the stability of complex systems. Variation in component
response rate becomes more likely to be critical for stability as system
size increases, and can ultimately increase the overall probability that
system stability is observed above that predicted by
May's\textsuperscript{\protect\hyperlink{ref-May1972}{1}} classically
derived \(\sigma \sqrt{SC}\) criterion. The logic outlined here is
general, and potentially applies to any complex system in which
individual system components can vary in their reaction rates to system
perturbation.

It is important to emphasise that variation in component response rate
is not stabilising per se; that is, adding variation in component
response rate to a particular system does not necessarily increase the
probability that the system will be stable. Rather, systems that are
observed to be stable are more likely to have varying component response
rates, and for this variation to be critical to their stability (Fig.
3). This is caused by the shift to a non-uniform distribution of
eigenvalues that occurs by introducing \(Var(\gamma)\) (Fig. 1b, 2b),
which can sometimes cause all of the real components of the eigenvalues
of the community matrix \(\mathbf{M}\) to become negative, but might
also increase the real components of eigenvalues. The mathematics
underlying this shift in eigenvalue distribution has been
investigated\textsuperscript{\protect\hyperlink{ref-Ahmadian2015}{16}}
and recently applied to questions concerning species density and
feasibility\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{17},\protect\hyperlink{ref-Stone2017}{18}},
but has not been interpreted as rates of response of individual system
components to perturbation.

@@

Interestingly, while complex systems were more likely to be stable given
variation in component response rate, they were not more likely to be
feasible, meaning that stability was not increased when component values
were also restricted to being positive at equilibrium. Feasibility is
important to consider, particularly for the study of ecological networks
of
species\textsuperscript{\protect\hyperlink{ref-Grilli2017}{5},\protect\hyperlink{ref-Stone2017}{18},\protect\hyperlink{ref-Dougoud2018}{20},\protect\hyperlink{ref-Servan2018}{22}}
because population densities cannot realistically be negative. My
results therefore suggest that variation in the rate of population
responses to perturbation (e.g., due to differences in generation time
among species) is unlikely to be critical to the feasibility of purely
multi-species interaction networks (see also Supplementary Information).
Nevertheless, ecological interactions do not exist in isolation in
empirical
systems\textsuperscript{\protect\hyperlink{ref-Patel2018}{15}}, but
instead interact with evolutionary, abiotic, or social-economic systems.
The relevance of component response rate for complex system stability
should therefore not be ignored in the broader context of ecological
communities.

@@

\textbf{Component response rate variation (\(\mathbf{\gamma}\))}. In a
synthesis of eco-evolutionary feedbacks on community stability, Patel et
al. model a system that includes a vector of potentially changing
species densities (\(\mathbf{N}\)) and a vector of potentially evolving
traits
(\(\mathbf{x}\))\textsuperscript{\protect\hyperlink{ref-Patel2018}{15}}.
For any species \(i\) or trait \(j\), change in species density
(\(N_{i}\)) or trait value (\(x_{j}\)) with time (\(t\)) is a function
of the vectors \(\mathbf{N}\) and \(\mathbf{x}\),
